---
title: "Multi-omics data preprocessing"
output: html_notebook
---

# Scope

This notebook documents the preprocessing used to obtain the complete
multi-omics matrix `mat_purity_mut` and the stratified training and test sets.
The Drews compendium is shown as the general example because the same workflow
was applied to Steele and Tao, starting from their respective signature
matrices.

The final `mat_purity_mut` matrices will be distributed through Zenodo. Raw
TCGA data and external signature source files are not redistributed in this
repository.

# Input paths

The original TCGA and signature inputs are available from the studies cited in
the manuscript. The paths below are placeholders and should be adapted to the
local locations of the downloaded files. Because several preprocessing steps
are computationally intensive, users interested in reproducing the downstream
analyses can start directly from the final matrices distributed with the
associated Zenodo record.

```{r, eval=FALSE}
input_dir <- "/path/to/downloaded_inputs"

post_normalization_file <- file.path(
  input_dir,
  "exp_meth_post_normalization.RData"
)
mutation_file <- "../data/processed/mut_finale.RData"
age_file <- file.path(input_dir, "mmc1.xlsx")
ucec_age_file <- file.path(
  input_dir,
  "TCGA-UCEC_clinical_from_XML.csv"
)

drews_signature_file <- file.path(
  input_dir,
  "Signature_Compendium_v5_Cosine-0.74_Activities_THRESH95_NAMESAPRIL21.rds"
)
steele_signature_file <- file.path(
  input_dir,
  "Steele_signature_attributions.xlsx"
)
steele_metadata_file <- file.path(
  input_dir,
  "Steele_ASCAT_metadata.xlsx"
)
tao_signature_file <- file.path(
  input_dir,
  "tcga_cn_sigs_CN176_activity.rds"
)
tao_type_file <- file.path(input_dir, "TCGA_type_info.rds")
```

The object `ordinata` loaded from `exp_meth_post_normalization.RData` is
the harmonized pan-cancer expression and methylation matrix. The methylation
normalization and batch-effect assessment underlying this matrix are
documented in `05_methylation_batch_effect.Rmd`.

The age table represented by `mmc1.xlsx` is the supplementary clinical table
from the source study specified in the manuscript. File names may differ after
download and do not need to match these placeholders.

# R packages
```{r, eval=FALSE}
library(TCGAbiolinks)
library(TCGAutils)
library(curatedTCGAData)
library(MultiAssayExperiment)
library(caret)
library(dplyr)
library(readr)
library(readxl)
library(usefun)
library(KEGGREST)
library(org.Hs.eg.db)
```

# Compendium-specific input

Only the signature source and the associated sample annotations change among
the three compendia.

| Compendium | Target shown in preprocessing | Signature source |
| --- | --- | --- |
| Drews | CX1 | Drews signature activity matrix |
| Steele | CN1 | Steele pan-cancer signature attribution matrix |
| Tao | Sig1 | Tao relative signature activity matrix |

## Drews
```{r, eval=FALSE}
exp<-readRDS(drews_signature_file)
load(post_normalization_file)

rownames(exp)<-substr(rownames(exp),1, 12)
exp_INT<-intersect(rownames(exp), rownames(ordinata))
un_tumor_finale<-ordinata[exp_INT, , drop = FALSE]
```

## Steele
```{r, eval=FALSE}
steel<- read_excel(
  steele_signature_file,
  sheet = "Pancan sig attributions"
)
steel<- data.frame(steel, row.names = 1)
steel<-steel[order(row.names(steel)),]
steel[sapply(steel, is.character)] <-
  lapply(steel[sapply(steel, is.character)], as.numeric)
rownames(steel)<-substr(rownames(steel),1, 12)

metadata1<-read_excel(
  steele_metadata_file,
  sheet = "TCGA ASCAT fits"
)
meta<-data.frame(metadata1, row.names = 1)
rownames(meta)<-substr(rownames(meta),1, 12)
meta<-meta[order(row.names(meta)),]

steel <- steel %>%
  mutate(purity = meta$purity) %>%
  mutate(cancer_type = meta$cancer_type) %>%
  relocate(purity, .after = cancer_type)

exp_INT<-intersect(rownames(steel), rownames(ordinata))
un_tumor_finale<-ordinata[exp_INT, , drop = FALSE]
```

## Tao
```{r, eval=FALSE}
tcga_cn_sigs_CN176_activity <- readRDS(
  tao_signature_file
)
TCGA<-tcga_cn_sigs_CN176_activity$relative
TCGA <- data.frame(TCGA, row.names = 1)

TCGA_type_info <- readRDS(tao_type_file)
prova<-TCGA[grep("01$|03$", rownames(TCGA)),]
rownames(prova)<-substr(rownames(prova),1, 12)

tao_ct<-data.frame(TCGA_type_info, row.names = 1)
tao_ct$samples<-rownames(tao_ct)
tao_ct<-as.data.frame(tao_ct[order(row.names(tao_ct)),])
tao_ct<-tao_ct[grep("01$|03$", rownames(tao_ct)),]
rownames(tao_ct)<-substr(rownames(tao_ct), 1, 12)
flag<-prova[rownames(prova) %in% rownames(tao_ct),]

tao<-cbind(tao_ct[,1], flag)
colnames(tao)[1]<-"cancer_type"

exp_INT<-intersect(rownames(tao), rownames(ordinata))
un_tumor_finale<-ordinata[exp_INT, , drop = FALSE]
```

# Expression and methylation filtering

The pan-cancer expression and methylation matrices were first restricted to
samples represented in the selected signature compendium. Features in the
lowest 20% of the coefficient-of-variation distribution were removed
separately for expression and methylation.

```{r, eval=FALSE}
un_Tumor_scale<-scale(un_tumor_finale, center =FALSE)

CV1<-sapply(
  as.data.frame(un_Tumor_scale),
  function(un_Tumor_scale)
    sd(un_Tumor_scale, na.rm=TRUE) /
    abs(mean(un_Tumor_scale, na.rm= TRUE))
)
C_V1<-as.matrix(CV1)

# Drews expression features
b<-as.matrix(C_V1[1:16400,])
b<-as.data.frame(b)
o1<-as.matrix(arrange(b, b$V1))
fd1<-as.data.frame(o1[3281:16400,])
colnames(fd1)[1] <- "V1"

# Drews methylation features
c<-as.matrix(C_V1[16401:30083,])
c<-as.data.frame(c)
o<-as.matrix(arrange(c, c$V1))
fd<-as.data.frame(o[2738:13683,])
colnames(fd)[1] <- "V1"

fd_exp<-as.matrix(fd1)
ordi_exp<-as.matrix(fd_exp[order(row.names(fd_exp)), ])
fd_meth<-as.matrix(fd)
ordi_meth<-as.matrix(fd_meth[order(row.names(fd_meth)), ])

test_exp<-un_tumor_finale[,1:16400]
test_meth<-un_tumor_finale[, 16401:30083]
prova_exp<-test_exp[,colnames(test_exp) %in% colnames(t(ordi_exp))]
prova_meth<-test_meth[,colnames(test_meth) %in% colnames(t(ordi_meth))]
un_Tumor_CV<-cbind(prova_exp, prova_meth)
```

For Steele and Tao, the corresponding original indices were:
```{r, eval=FALSE}
b<-as.matrix(C_V1[1:15976,])
b<-as.data.frame(b)
o1<-as.matrix(arrange(b, b$V1))
fd1<-as.data.frame(o1[3196:15976,])

c<-as.matrix(C_V1[15977:29650,])
c<-as.data.frame(c)
o<-as.matrix(arrange(c, c$V1))
fd<-as.data.frame(o[2736:13674,])

test_exp<-un_tumor_finale[,1:15976]
test_meth<-un_tumor_finale[,15977:29650]
```

# Near-zero variance and correlation filtering
```{r, eval=FALSE}
a<-as.character(colnames(un_Tumor_CV))
b<-make.unique(a)
colnames(un_Tumor_CV)<-b

nzv<-nearZeroVar(
  un_Tumor_CV,
  uniqueCut = 40,
  saveMetrics = TRUE,
  names = TRUE
)

nzv.1<-nearZeroVar(
  un_Tumor_CV,
  uniqueCut = 40,
  saveMetrics = FALSE
)

M<-cor(un_Tumor_CV, use = "pairwise.complete.obs")
flag<-findCorrelation(M, cutoff = .90)
un_Tumor_CV_f<-un_Tumor_CV[, -flag]
```

# Signature and clinical variables

The selected signature was placed at the beginning of the matrix. Cancer type,
age, and purity were then added from the corresponding TCGA and ASCAT
annotations.

## Drews example
```{r, eval=FALSE}
s_un_Tumor<-scale(un_Tumor_CV_f, center =FALSE)

exp_IN<-exp[exp_INT, , drop = FALSE]
un_tumor_finale<-as.data.frame(cbind(exp_IN[, 1], s_un_Tumor))
colnames(un_tumor_finale)[1]<-"CX1"
un_tumor_finale<-
  un_tumor_finale[, colSums(is.na(un_tumor_finale))==0]

metadata<-Metadata_TCGA_ASCAT_penalty70
p_ct<-metadata[, c(2,2,3,8)]
dup<-!duplicated(p_ct[, 1])
flag<-p_ct[dup,]
p_ct<-data.frame(flag, row.names = 2)
p_ct<-p_ct[exp_INT, , drop = FALSE]
mat_p_ct<-cbind(p_ct, un_tumor_finale)

c<-as.data.frame(cbind(
  colData(primaryTumors)[, getClinicalNames("ACC")][, 1:2]@rownames,
  colData(primaryTumors)[, getClinicalNames("ACC")][, 1:2]$years_to_birth
))
colnames(c)[c(1,2)]<- c("patient", "age")
rownames(c)<-c[,1]
c<-c[order(rownames(c)), ]

c_sorf<-intersect(rownames(c), rownames(un_Tumor_CV_f))
c_sorf<-c[c_sorf, , drop = FALSE]
mat_finale<-cbind(c_sorf, mat_p_ct[,-1])
```

### Drews-specific recovery of missing UCEC ages

The original Drews workflow included an additional step to recover missing
UCEC ages from the TCGA clinical XML data before complete cases were selected.

```{r, eval=FALSE}
TCGA_UCEC_clinical_from_XML <- read_csv(ucec_age_file)
a<-as.data.frame(cbind(
  TCGA_UCEC_clinical_from_XML$days_to_birth,
  TCGA_UCEC_clinical_from_XML$bcr_patient_barcode
))
a<-a[rowSums(is.na(a)) ==0,]
colnames(a)<-c("age", "patient")
a<-unique(a)
eta<-data.frame(a, row.names = 2)

eta[sapply(eta, is.character)] <-
  lapply(eta[sapply(eta, is.character)], as.integer)
eta_finale<-as.data.frame(
  sapply(eta, function(x) floor(abs(x) /365.25, 0))
)
rownames(eta_finale)<-as.vector(rownames(eta))
eta_finale<-cbind(eta_finale, rownames(eta))
ordi_eta<-as.data.frame(
  eta_finale[order(row.names(eta_finale)), ]
)
colnames(ordi_eta)[2]<-"patient"

vet<-intersect(rownames(ordi_eta), rownames(mat_finale))
visual<-ordi_eta[vet, , drop = FALSE]
m_visual<-mat_finale[vet, , drop = FALSE]

m_all_ucec<-cbind(visual,m_visual[,-c(1,2)])
m_visual_no_ucec<-
  mat_finale[!rownames(mat_finale) %in% rownames(ordi_eta),]

m_finale_ucec<-rbind(m_visual_no_ucec, m_all_ucec)
m_finale_ucec_ordinata<-as.data.frame(
  m_finale_ucec[order(row.names(m_finale_ucec)), ]
)
```

## Steele-specific clinical variables

Steele used the ASCAT cancer type and purity supplied with that compendium.
Three samples missing from the age table were added explicitly in the original
workflow.

```{r, eval=FALSE}
exp_IN<-steel[exp_INT, , drop = FALSE]
un_tumor_finale<-as.data.frame(
  cbind(CN1 = exp_IN[, 1], un_Tumor_CV_f)
)
un_tumor_finale<-
  un_tumor_finale[, colSums(is.na(un_tumor_finale))==0]

un_tumor_finale <- un_tumor_finale %>%
  mutate(cancer_type = exp_IN$cancer_type) %>%
  mutate(purity = exp_IN$purity) %>%
  relocate(cancer_type, .after = CN1) %>%
  relocate(purity, .after = cancer_type)

mmc1 <- read_excel(age_file)
tt<-mmc1[, c(2:4)]
tt<-data.frame(tt, row.names = 1)
tt1<-tt[order(rownames(tt)),]
tmp1<-intersect(rownames(tt1), rownames(un_tumor_finale))
tt2<-tt1[tmp1, , drop = FALSE]

tt3 <- data.frame(
  type = c("GBM","READ", "LGG"),
  age_at_initial_pathologic_diagnosis = c(NA, 71, NA),
  row.names = c(
    "TCGA-28-2510",
    "TCGA-F5-6810",
    "TCGA-R8-A6YH"
  )
)
tt3<-rbind(tt2, tt3)
tt3<-tt3[order(rownames(tt3)),]

mat_finale<-cbind(age = tt3[, 2], un_tumor_finale)
mat_finale <- mat_finale %>%
  relocate(age, .before = purity)
```

## Tao-specific clinical variables

Tao used its own tumor-type annotation, while purity was matched from the
Steele ASCAT metadata and age was matched from the TCGA clinical table.

```{r, eval=FALSE}
exp_IN<-tao[exp_INT, , drop = FALSE]
un_tumor_finale<-as.data.frame(
  cbind(Sig1 = exp_IN$Sig1, un_Tumor_CV_f)
)
un_tumor_finale<-
  un_tumor_finale[, colSums(is.na(un_tumor_finale))==0]

un_tumor_finale<- un_tumor_finale %>%
  mutate(cancer_type = exp_IN$cancer_type) %>%
  relocate(cancer_type, .before = Sig1)

metadata1<-read_excel(
  steele_metadata_file,
  sheet = "TCGA ASCAT fits"
)
meta<-data.frame(metadata1, row.names = 1)
rownames(meta)<-substr(rownames(meta),1, 12)
meta<-meta[order(row.names(meta)),]
meta<-meta[exp_INT, , drop = FALSE]

un_tumor_finale<- un_tumor_finale %>%
  mutate(purity = meta$purity) %>%
  relocate(purity, .after = Sig1)

mmc1 <- read_excel(age_file)
tt<-mmc1[, c(2:4)]
tt<-data.frame(tt, row.names = 1)
tt1<-tt[order(rownames(tt)),]
tmp1<-intersect(rownames(tt1), rownames(un_tumor_finale))
tt2<-tt1[tmp1, , drop = FALSE]

mat_finale<-cbind(age = tt2[, 2], un_tumor_finale)
mat_finale <- mat_finale %>%
  relocate(age, .after = Sig1)
```

These compendium-specific steps are retained separately because they alter
sample inclusion and the positions of the identifier and clinical columns used
in the final matrix.

# Mutation and pathway features

The mutation matrix was reduced by replacing genes represented in the selected
cancer-related and KEGG pathways with pathway-level mutation counts. Genes not
assigned to these pathways were retained individually.

```{r, eval=FALSE}
load(mutation_file)
```

The objects `pathway` and `names_geneSymbol` contain the pathway-to-gene
mappings used in the original analysis.

```{r, eval=FALSE}
mut<-as.data.frame(do.call(rbind, data_mut_finale))
rownames(mut) <- sub(".*\\.", "", rownames(mut))
df<-mut[order(row.names(mut)), ]
t<-na.omit(mat_finale)
tmp<-intersect(rownames(df), rownames(t))
df<-df[tmp, , drop = FALSE]

keep_cols<-list()
for (i in names(pathway)) {
  genes <- pathway[[i]]
  for(j in genes){
    matching_cols <- grep(
      paste0("^", j, "$"),
      colnames(df),
      value = TRUE
    )
    keep_cols[[names(pathway[i])]] <-
      unique(c(keep_cols[[i]], matching_cols))
  }
}

path_df<-list()
for (i in names(pathway)) {
  path_df[[names(pathway[i])]] <- df[, keep_cols[[i]]]
}

genes_to_drop<-colnames(do.call(cbind,path_df))
genes_to_drop<- sub(".*\\.", "", genes_to_drop)
tmp<-intersect(colnames(df), genes_to_drop)

df_mod<-df[, tmp, drop = FALSE]
df_finale <- data.frame(
  matrix(nrow = nrow(df), ncol = length(pathway))
)
rownames(df_finale)<-rownames(df_mod)
colnames(df_finale)<-names(pathway)

for (i in 1:length(pathway)) {
  temp_df <- df_mod %>%
    mutate(!!names(pathway[i]) := rowSums(path_df[[i]])) %>%
    relocate(!!names(pathway[i]))
  df_finale[, i]<-temp_df[,1]
}

keep_cols <- list()
for (i in names(names_geneSymbol)) {
  genes <- names_geneSymbol[[i]]
  for (j in genes) {
    matching_cols <- grep(
      paste0("^", j, "$"),
      colnames(df),
      value = TRUE
    )
    if (length(matching_cols) == 0) {
      gene_pattern <- substr(j, 1, nchar(j) - 1)
      matching_cols <- grep(
        paste0("^", gene_pattern, "$"),
        colnames(df),
        value = TRUE
      )
    }
    keep_cols[[i]] <- unique(c(keep_cols[[i]], matching_cols))
  }
}

path_df<-list()
for (i in names(names_geneSymbol)) {
  path_df[[names(names_geneSymbol[i])]] <- df[, keep_cols[[i]]]
}

genes_to_drop<-colnames(do.call(cbind,path_df))
genes_to_drop<- sub(".*\\.", "", genes_to_drop)
tmp<-intersect(colnames(df), genes_to_drop)

df_mod1<-df[, tmp, drop = FALSE]
df_finale1 <- data.frame(
  matrix(nrow = nrow(df), ncol = length(names_geneSymbol))
)
rownames(df_finale1)<-rownames(df_mod1)
colnames(df_finale1)<-names(names_geneSymbol)

for (i in 1:length(names_geneSymbol)) {
  temp_df <- df_mod1 %>%
    mutate(
      !!names(names_geneSymbol[i]) := rowSums(path_df[[i]])
    ) %>%
    relocate(!!names(names_geneSymbol[i]))
  df_finale1[, i]<-temp_df[,1]
}

df_finale_unito<-cbind(df_finale, df_finale1)
drop_col<-unique(c(colnames(df_mod), colnames(df_mod1)))
tt<-df[, !(colnames(df) %in% drop_col), drop = FALSE]
mut_path<-cbind(tt, df_finale_unito)
```

# Final multi-omics matrix

## Drews
```{r, eval=FALSE}
t<-na.omit(m_finale_ucec_ordinata)
prova<-t[,-c(2,3,5)]
prova<-cbind(prova, mut_path[13343:length(mut_path)])
prova1<-scale(prova, center = FALSE)
mat_purity_mut<-cbind(t[, c(2,3,5)], prova1)
mat_purity_mut<-cbind(
  mat_purity_mut,
  mut_path[, -c(1:2, 13343:length(mut_path)), drop = FALSE]
)

a<-as.character(colnames(mat_purity_mut))
b<-make.unique(a)
colnames(mat_purity_mut)<-b
```

## Steele and Tao
```{r, eval=FALSE}
t<-na.omit(mat_finale)
prova<-t[,-c(1,2)]
prova<-cbind(prova, mut_path[13343:length(mut_path)])
prova1<-scale(prova, center = FALSE)
mat_purity_mut<-cbind(t[, c(1,2)], prova1)
mat_purity_mut<-cbind(
  mat_purity_mut,
  mut_path[, -c(1, 13343:length(mut_path)), drop = FALSE]
)

a<-as.character(colnames(mat_purity_mut))
b<-make.unique(a)
colnames(mat_purity_mut)<-b

mat_purity_mut <-mat_purity_mut %>%
  relocate(Tumor_Sample_Barcode, .before = cancer_type)
```

# Alternative target signatures

The three distributed final matrices contain the first signature used during
preprocessing:

| Matrix | Included target column |
| --- | --- |
| `mat_purity_mut_drews` | `CX1` |
| `mat_purity_mut_steele` | `CN1` |
| `mat_purity_mut_tao` | `Sig1` |

```{r, eval=FALSE}
load("/path/to/final_matrices.RData")
```

The complete signature matrices are stored in
`data/processed/signature_exploration_data.RData` as `drew`, `steele`, and
`tao`. To model another signature, the original target column is replaced by
the desired signature after matching samples by their row names. All
multi-omics predictors remain unchanged.

```{r, eval=FALSE}
load("../data/processed/signature_exploration_data.RData")

# Example: replace CX1 with CX2 in the Drews matrix
sample_ids<-rownames(mat_purity_mut_drews)
stopifnot(all(sample_ids %in% rownames(drew)))

mat_purity_mut_drews$CX1<-drew[sample_ids, "CX2"]
colnames(mat_purity_mut_drews)[
  colnames(mat_purity_mut_drews) == "CX1"
]<-"CX2"
```

The same replacement is applied to Steele (`CN1`) and Tao (`Sig1`). For the
models reported in the manuscript, the alternative targets were:

```text
Drews:  CX2, CX3, CX5
Steele: CN2, CN9, CN17
Tao:    Sig2, Sig5, Sig7
```

The saved train/test sample identifiers should then be used to retain the
original split for every target signature.

# Stratified training and test split
An 80/20 split stratified by cancer type was used. The split was generated
independently for each compendium.

## Drews
```{r, eval=FALSE}
set.seed(1234)
fract<- mat_purity_mut_drews %>%
  group_by(cancer_type) %>%
  sample_frac(.80)

ordi_train<-data.frame(fract, row.names = 1)
tmp<-outersect(rownames(ordi_train), mat_purity_mut$patient)
ordi_test<-mat_purity_mut[tmp, , drop = FALSE]
ordi_test<-data.frame(ordi_test, row.names = 1)
```

## Steele
```{r, eval=FALSE}
set.seed(1234)
fract<- mat_purity_mut_steele %>%
  group_by(cancer_type) %>%
  sample_frac(.80)

ordi_train<-data.frame(fract, row.names = 2)
tmp<-outersect(rownames(ordi_train), rownames(mat_purity_mut))
ordi_test<-mat_purity_mut[tmp, , drop = FALSE]
ordi_test<-data.frame(ordi_test, row.names = 2)
```

## Tao
```{r, eval=FALSE}
set.seed(1234)
fract<- mat_purity_mut_tao %>%
  group_by(cancer_type) %>%
  sample_frac(.80)

ordi_train<-data.frame(fract, row.names = 1)
tmp<-outersect(rownames(ordi_train), rownames(mat_purity_mut))
ordi_test<-mat_purity_mut[tmp, , drop = FALSE]
ordi_test<-data.frame(ordi_test, row.names = 1)
```

# Distributed checkpoints

The computationally expensive correlation filtering and pathway construction
do not need to be repeated to reproduce the modeling analyses. The
accompanying Zenodo record should therefore contain one final matrix per
compendium:

```text
mat_purity_mut_drews.RData
mat_purity_mut_steele.RData
mat_purity_mut_tao.RData
```

Each file should contain the appropriately named `mat_purity_mut_*` object and
the original train/test sample identifiers. The matrices initially contain
only `CX1`, `CN1`, or `Sig1`; the other target signatures are obtained from
`signature_exploration_data.RData` as shown above. These inputs allow users to
reconstruct the final modeling matrices for any target signature or cluster
classification. The approximately 4 GB post-normalization object is not
included in GitHub.
