---
title: "Copy number signature exploration"
output: html_notebook
---

# Scope

This notebook generates Figures 2-4 and Supplementary Figure 1. It compares
the distributions, principal components, and agreement of copy number
signatures from the Drews, Steele, and Tao compendia.

# R packages and data
```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(reshape2)
library(ggplot2)
library(ggpubr)
library(factoextra)
library(pheatmap)
library(RColorBrewer)

load("../data/processed/signature_exploration_data.RData")

data<-list(
  Drews = drews,
  Steele = steele,
  Tao = tao
)
```
The input file contains the three complete signature matrices, the matrices
restricted to common samples, and the TCGA tissue source site table.

# Figure 2

## Signature distributions
```{r}
a<-melt(drews[,-1], variable.name = "signature")
drews_p<-ggplot(a, aes(x = signature, y = value, fill = signature)) +
  geom_boxplot(outlier.size = .2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank()) +
  guides(fill = "none")

a1<-melt(steele[,-1], variable.name = "signature")
steele_p<-ggplot(a1, aes(x = signature, y = value, fill = signature)) +
  geom_boxplot(outlier.size = .2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank()) +
  guides(fill = "none")

a1<-melt(tao[,-1], variable.name = "signature")
tao_p<-ggplot(a1, aes(x = signature, y = value, fill = signature)) +
  geom_boxplot(outlier.size = .2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank()) +
  guides(fill = "none")
```

## PCA loadings and linear models
```{r}
regression_feature_selection <- function(data_sig) {
  codici<-substr(rownames(data_sig), 6,7)
  flag<- data_sig %>%
    mutate(TSS_Code = factor(codici, levels = unique(codici)),
           sample = rownames(data_sig))
  tmp<-merge(tissueSourceSite, flag, by = "TSS_Code")
  merge_df<-tmp[,-c(3:4)]
  rownames(merge_df)<-merge_df$sample
  colnames(merge_df)[2] <- "source_site"
  merge_df<-merge_df[, -c(1,length(merge_df))]
  return(merge_df)
}

df_regresion<-list()
for (i in seq_along(data)) {
  df_regresion[[names(data[i])]]<- regression_feature_selection(data[[i]])
}

plot_list <- list()
for (i in seq_along(data)) {
  pca_result<-prcomp(data[[i]][, -1], scale = FALSE)
  plot_list[[i]]<-fviz_pca_var(
    pca_result,
    col.var = "contrib",
    gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
  )

  pca_result_reg<-prcomp(df_regresion[[i]][, -c(1:2)], scale = TRUE)
  pca_data <- as.data.frame(pca_result_reg$x[, 1:5])
  pca_data$source_site <- df_regresion[[i]]$source_site
  pca_data$cancer_type <- df_regresion[[i]]$cancer_type

  model_metrics <- lapply(1:5, function(i) {
    model <- lm(get(paste0("PC", i)) ~ source_site, data = pca_data)
    summary <- summary(model)
    data.frame(
      PC = paste0("PC", i),
      R_squared = summary$r.squared,
      F_statistic = summary$fstatistic[1],
      RSE = summary$sigma
    )
  }) %>%
    bind_rows()

  model_metrics1 <- lapply(1:5, function(i) {
    model <- lm(get(paste0("PC", i)) ~ cancer_type, data = pca_data)
    summary <- summary(model)
    data.frame(
      PC = paste0("PC", i),
      R_squared = summary$r.squared,
      F_statistic = summary$fstatistic[1],
      RSE = summary$sigma
    )
  }) %>%
    bind_rows()

  plot_list[[i+3]]<-ggplot(model_metrics, aes(x = PC, y = R_squared)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(y = "Value", fill = "Variable") +
    ylab("R_squared TSS") +
    xlab(NULL) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_y_continuous(limits = c(0, .35))

  plot_list[[i+6]]<-ggplot(model_metrics1, aes(x = PC, y = R_squared)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(x = "PC", y = "Value", fill = "Variable") +
    ylab("R_squared Tumor type") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_y_continuous(limits = c(0, .35))
}
```

```{r figure-2, fig.width=14, fig.height=18}
plotlist <- c(
  list(drews = drews_p, steele = steele_p, tao = tao_p),
  plot_list
)

figure2<-ggarrange(
  plotlist = plotlist,
  heights = c(1.75, 1.75, 0.75, 0.75),
  labels = "auto",
  vjust = 1.2,
  nrow = 4,
  ncol = 3
)

figure2
```

# Figure 3 and Supplementary Figure 1
```{r}
data4 <- list(
  Drews = cor(data[[1]][,-1]),
  Drews_Tao = cor(drews_tao[,-1], tao_drews[,-1]),
  Steele = cor(data[[2]][,-1]),
  Drews_Steele = cor(
    drews_tao[,-1],
    steele_drews[,-1],
    use = "complete.obs"
  ),
  Tao = cor(data[[3]][,-1]),
  Tao_Steele = cor(
    tao_drews[,-1],
    steele_drews[,-1],
    use = "complete.obs"
  )
)

data6 <- list(
  Drews = cor(data[[1]][,c(2:4, 6)]),
  Drews_Tao = cor(drews_tao[,c(2:4, 6)], tao_drews[,c(2:3, 6,8)]),
  Steele = cor(data[[2]][,c(2:3, 10, 18)]),
  Drews_Steele = cor(
    drews_tao[,c(2:4, 6)],
    steele_drews[,c(2:3, 10, 18)],
    use = "complete.obs"
  ),
  Tao = cor(data[[3]][,c(2:3, 6,8)]),
  Tao_Steele = cor(
    tao_drews[,c(2:3, 6,8)],
    steele_drews[,c(2:3, 10, 18)],
    use = "complete.obs"
  )
)

pheatmap_colors <- colorRampPalette(
  rev(brewer.pal(n = 7, name = "RdYlBu"))
)(100)
```

```{r figure-3, fig.width=12, fig.height=16}
plot_list <- list()

for (i in seq_along(data4)) {
  corr_df <- melt(data4[[i]])

  p <- ggplot(corr_df, aes(x = Var1, y = Var2, fill = value)) +
    geom_tile(color = "black") +
    geom_text(aes(label = round(value, 2)), size = 2) +
    scale_fill_gradientn(
      colors = pheatmap_colors,
      limits = c(-0.8, 1),
      name = "Correlation"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.text.y = element_text(hjust = 1.7),
      axis.title = element_blank()
    )

  plot_list[[i]] <- p
}

corr_plot <- ggarrange(
  plotlist = plot_list,
  ncol = 2,
  nrow = 3,
  labels = "auto",
  font.label = list(size = 8),
  common.legend = TRUE,
  legend = "bottom"
)

corr_plot
```

```{r supplementary-figure-1, fig.width=10, fig.height=12}
plot_list <- list()

for (i in seq_along(data6)) {
  corr_df <- melt(data6[[i]])

  p <- ggplot(corr_df, aes(x = Var1, y = Var2, fill = value)) +
    geom_tile(color = "black") +
    geom_text(aes(label = round(value, 2)), size = 3) +
    scale_fill_gradientn(
      colors = pheatmap_colors,
      limits = c(-1, 1),
      name = "Correlation"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.text.y = element_text(hjust = 1),
      axis.title = element_blank()
    )

  plot_list[[i]] <- p
}

corr_plot_active <- ggarrange(
  plotlist = plot_list,
  ncol = 2,
  nrow = 3,
  labels = "auto",
  font.label = list(size = 8),
  common.legend = TRUE,
  legend = "bottom"
)

corr_plot_active
```

# Figure 4
```{r jaccard-calculation, fig.show="hide"}
replace_median<-function(x) {
  median_val<-median(x, na.rm = TRUE)
  if (median_val == 0) {
    return(ifelse(x > median_val, 1, 0))
  } else {
    return(ifelse(x >= median_val, 1, 0))
  }
}

drews_dico<- drews[,-1] %>%
  mutate(across(everything(), replace_median))
drews_dico<-drews_dico[rownames(drews_tao), , drop = FALSE]

tao_dico<- tao[,-1] %>%
  mutate(across(everything(), replace_median))
tao_dico<-tao_dico[rownames(tao_drews), , drop = FALSE]

steele_dico<- steele[,-1] %>%
  mutate(across(everything(), replace_median))
steele_dico<-steele_dico[rownames(steele_drews), , drop = FALSE]

complete_samples<-complete.cases(steele_dico)
drews_dico<-drews_dico[complete_samples, , drop = FALSE]
tao_dico<-tao_dico[complete_samples, , drop = FALSE]
steele_dico<-steele_dico[complete_samples, , drop = FALSE]

flag<-cbind(tao_dico, drews_dico)
flag2<-cbind(flag, steele_dico)
x<-flag2

jaccard_mat<-function(x) {
  Jc_matrix<-matrix(NA, nrow = ncol(x), ncol = ncol(x))
  for (i in 1:ncol(x)) {
    for (j in 1:ncol(x)) {
      if (i != j) {
        true<-sum(x[,i] == x[,j])
        tot<-2*nrow(x) - true
        Jc_matrix[i, j]<- true/tot
      } else {
        Jc_matrix[i, j]<- NA
      }
    }
  }
  colnames(Jc_matrix)<-colnames(x)
  rownames(Jc_matrix)<-colnames(x)
  return(Jc_matrix)
}

b<-jaccard_mat(x)

phe<-pheatmap(flag2, show_rownames = FALSE)
column_order_indices<-phe[["tree_col"]][["order"]]
column_order_names <- colnames(flag2)[column_order_indices]

plot_heatmap <- function(risultati, col_labels) {
  risultati <- risultati[col_labels, col_labels, drop = FALSE]
  matrice_long <- reshape2::melt(
    risultati,
    varnames = c("Var1", "Var2")
  )

  ggplot(matrice_long, aes(x = Var1, y = Var2, fill = value)) +
    geom_tile(
      color = "grey25",
      lwd = .1,
      linetype = 1
    ) +
    scale_fill_gradientn(
      colors = c("white", "white", "lightblue", "darkblue", "darkblue"),
      values = scales::rescale(c(0, 0.25, 0.5, .75, 1)),
      breaks = c(0, 0.25, 0.5, 0.75, 1),
      limits = c(0, 1),
      na.value = "grey50"
    ) +
    labs(x = "", y = "", fill = "Value") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 50, hjust = 1, size = 7),
      axis.text.y = element_text(size = 8),
      legend.position = "none"
    ) +
    coord_fixed()
}

all<-plot_heatmap(b, column_order_names)

y<-x[, c(1,2,5,7,21:23,25,38:39,46,54)]
phe<-pheatmap(y, show_rownames = FALSE)
column_order_indices<-phe[["tree_col"]][["order"]]
column_order_names <- colnames(y)[column_order_indices]

all_active<-plot_heatmap(b, column_order_names) +
  theme(legend.position = "right")
```

```{r figure-4, fig.width=14, fig.height=7}
jc_list<-list(all, all_active)

figure4<-ggarrange(
  plotlist = jc_list,
  labels = "auto",
  vjust = 2.5,
  hjust = -2,
  widths = c(1.1, 0.9)
)

figure4
```
