---
title: "Copy number signature clustering"
output: html_notebook
---

# Scope

This notebook generates Figure 5 using the final signature matrices and cluster
classifications for the Drew, Steele, and Tao compendia. Clustering is not
recomputed in this notebook.

# R packages and data

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(pheatmap)
library(ggpubr)
library(usefun)

load("../data/processed/clustering_data.RData")
```

The input file contains `active_drew`, `active_steele`, `active_tao`,
`mc_drew_active_4`, `mc_steele_active_4`, and `mc_tao_active_4`.

# Signature heatmaps
```{r}
order.ind<-order(mc_drew_active_4$classification)
tmp<-active_drew[order.ind, -1]
annotation_col <- data.frame(
  Cluster = as.factor(mc_drew_active_4$classification[order.ind])
)
ann_colors = list(
  Cluster = c("1"="#72F59A", "2"="#F57D72","3"="#F7C073","4"="#8672F5"))
rownames(annotation_col) <- row.names(tmp)
drew_pheatmap<-pheatmap(tmp,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         annotation_row  = annotation_col,
         annotation_colors = ann_colors,
         cutree_rows = 4,
         show_rownames  = FALSE,
         angle_col = 90,
         width = 5,
         cellwidth = 15,
         height = 20)

order.ind<-order(mc_steele_active_4$classification)
tmp<-active_steele[order.ind, -1]
annotation_col <- data.frame(
  Cluster = as.factor(mc_steele_active_4$classification[order.ind])
)
rownames(annotation_col) <- row.names(tmp)
steele_pheatmap<-pheatmap(tmp,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         annotation_row  = annotation_col,
         annotation_colors = ann_colors,
         cutree_rows = 4,
         show_rownames  = FALSE,
         width = 5,
         angle_col = 90,
         cellwidth = 15,
         height = 20)

order.ind<-order(mc_tao_active_4$classification)
tmp<-active_tao[order.ind, -1]
annotation_col <- data.frame(
    Cluster = as.factor(mc_tao_active_4$classification[order.ind])
)
ann_colors = list(
    Cluster = c("1"="#72F59A", "2"="#F57D72", "3"="#F7C073"))
rownames(annotation_col) <- row.names(tmp)
tao_pheatmap<-pheatmap(tmp,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         annotation_row  = annotation_col,
         annotation_colors = ann_colors,
         cutree_rows = 4,
         show_rownames  = FALSE,
         width = 5,
         angle_col = 90,
         cellwidth = 15,
         height = 20)
```

# Signature distributions across clusters
```{r}
ann_colors = list(
  Cluster = c("1"="#72F59A", "2"="#F57D72","3"="#F7C073","4"="#8672F5")
)

# Function to create color scales based on cluster colors
create_color_scale <- function(base_color, n_colors) {
  colorRampPalette(c("white", base_color))(n_colors + 1)[-1]  # Remove white
}

# Function to create violin plot for any dataset
create_violin_plot <- function(df_long, title = "") {
  # Get unique variables and groups
  unique_vars <- unique(df_long$Variable)
  unique_groups <- unique(df_long$Group)
  n_vars <- length(unique_vars)
  
  # Create color mapping for each Variable-Group combination
  color_mapping <- c()
  for(group in unique_groups) {
    base_color <- ann_colors$Cluster[as.character(group)]
    group_colors <- create_color_scale(base_color, n_vars)
    
    for(i in seq_along(unique_vars)) {
      var <- unique_vars[i]
      key <- paste0(var, ".", group)
      color_mapping[key] <- group_colors[i]
    }
  }
  
  # Create fill variable for mapping
  df_long$fill_var <- interaction(df_long$Variable, df_long$Group)
  
  # Create the plot
  ggplot(df_long, aes(x = Variable, y = Value, fill = fill_var)) +
    geom_violin(trim = FALSE, scale = "width", color = "black", alpha = 0.8) +
    scale_fill_manual(values = color_mapping) +
    stat_summary(fun = mean, geom = "point", shape = 21, fill = "black", size = 2.5) +
    stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2, color = "black") +
    facet_wrap(~ Group, nrow = 1) +
    coord_cartesian(ylim = c(0, 1)) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) 
}

# Drew
df_drew <- as.data.frame(active_drew[,-1])
df_drew$Group <- mc_drew_active_4$classification
df_long_drew <- pivot_longer(df_drew, cols = starts_with("CX"), names_to = "Variable", values_to = "Value")

# Steele
df_steele <- as.data.frame(active_steele[,-1])
df_steele$Group <- mc_steele_active_4$classification
df_long_steele <- pivot_longer(df_steele, cols = starts_with("CN"), names_to = "Variable", values_to = "Value")

# Tao
df_tao <- as.data.frame(active_tao[,-1])
df_tao$Group <- mc_tao_active_4$classification
df_long_tao <- pivot_longer(df_tao, cols = starts_with("Sig"), names_to = "Variable", values_to = "Value")

drew_violin<-create_violin_plot(df_long_drew, "Drew Dataset")
steele_violin<-create_violin_plot(df_long_steele, "Steele Dataset")
tao_violin<-create_violin_plot(df_long_tao, "Tao Dataset")
```

# Tumor type composition
```{r}
# Function to create bar plot for cluster distribution by cancer type
create_cluster_barplot <- function(active_data, mc_result, order_by_cluster = 1, n_clusters = 4) {
  
  # Define cluster colors
  cluster_colors <- c("1"="#72F59A", "2"="#F57D72", "3"="#F7C073", "4"="#8672F5")
  
  # Create clustered data
  mydata_cl <- active_data %>%
    mutate(cluster = mc_result$classification)
  
  # Calculate aggregated data
  agg_data <- mydata_cl %>%
    group_by(cancer_type, cluster) %>%
    summarise(count = n(), .groups = 'drop') %>%
    group_by(cancer_type) %>%
    mutate(relative_freq = count / sum(count))
  
  # Order levels based on specified cluster
  order_levels <- agg_data %>%
    filter(cluster == order_by_cluster) %>%
    arrange(desc(relative_freq)) %>%   
    pull(cancer_type)
  
  # Get all cancer types
  all_cancers <- agg_data %>%
    distinct(cancer_type)
  
  # Complete ordering (handle missing cancer types)
  order_levels <- order_levels %>%
    c(outersect(as.character(.), all_cancers$cancer_type))
  
  # Apply factor ordering
  agg_data <- agg_data %>%
    mutate(cancer_type = factor(cancer_type, levels = order_levels))
  
  # Create plot
  ggplot(agg_data, aes(x = cancer_type, y = relative_freq)) +
    geom_bar(aes(fill = as.factor(cluster)), stat = "identity", position = "fill") +
    labs(y = "Relative Frequency", x = "Tumor Type", fill = "Cluster") +
    scale_fill_manual(values = cluster_colors[1:n_clusters]) +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
      axis.title = element_text(size = 7.5)
    )
}

drew_barplot <- create_cluster_barplot(
  active_data = active_drew,
  mc_result = mc_drew_active_4,
  order_by_cluster = 1,
  n_clusters = 4
)

steele_barplot <- create_cluster_barplot(
  active_data = active_steele,
  mc_result = mc_steele_active_4,
  order_by_cluster = 1,
  n_clusters = 4
)

tao_barplot <- create_cluster_barplot(
  active_data = active_tao,
  mc_result = mc_tao_active_4,
  order_by_cluster = 1,
  n_clusters = 3
)
```

# Figure 5
```{r}
plots_list<-list(drew_pheatmap$gtable, drew_violin, drew_barplot, 
                 steele_pheatmap$gtable, steele_violin, steele_barplot,
                 tao_pheatmap$gtable, tao_violin, tao_barplot)

ggarrange(plotlist = plots_list, labels = "AUTO", nrow = 3, ncol = 3,
          widths = c(0.6, 1.4, 1.2), heights = c(.9, 1, 1.1))
```
