---
title: "Survival analyses"
output: html_notebook
---

# Scope

This notebook documents the Kaplan-Meier and stratified Cox analyses performed
for the Drew, Steele, and Tao signature clusters. Overall survival (OS) and
progression-free interval (PFI) were analyzed separately.

The input `.RData` file contains the final `drew_f`, `steele_f`, and `tao_f`
survival datasets, the final `mc_drew_active_4`, `mc_steele_active_4`, and
`mc_tao_active_4` cluster objects, and the `purezza` table.

# R packages and data

```{r setup, message=FALSE, warning=FALSE}
library(survival)
library(survminer)
library(gtsummary)
library(dplyr)
library(ggplot2)

load("../data/processed/surv_data.RData")
```

# Kaplan-Meier analyses

## Drews
```{r}
a<-surv_cutpoint(drew_f,
              time = "OS.time",
              event = "OS",
              variables = c("CX1" , "CX2" , "CX3" , "CX5"),
              progressbar = TRUE
)
res.cat <- surv_categorize(a)
res.cat$cluster <- mc_drew_active_4$classification[rownames(res.cat)]

fit <- survfit(Surv(OS.time, OS) ~ cluster, data = res.cat)
drew_OS<-ggsurvplot(fit, data = res.cat, risk.table = TRUE, conf.int = FALSE, pval = TRUE, palette  = c("#72F59A", "#F57D72","#F7C073","#8672F5"), censor = FALSE)

a<-surv_cutpoint(drew_f,
              time = "PFI.time",
              event = "PFI",
              variables = c("CX1" , "CX2" , "CX3" , "CX5"),
              #minprop = 0.3,
              progressbar = TRUE
)
res.cat <- surv_categorize(a)
res.cat$cluster <- mc_drew_active_4$classification[rownames(res.cat)]

fit <- survfit(Surv(PFI.time, PFI) ~ cluster, data = res.cat)
drew_PFI<-ggsurvplot(fit, data = res.cat, risk.table = TRUE, conf.int = FALSE, pval = TRUE, palette  = c("#72F59A", "#F57D72","#F7C073","#8672F5"), censor = FALSE)
```

## Steele
```{r}
a<-surv_cutpoint(steele_f,
              time = "OS.time",
              event = "OS",
              variables = c("CN1" , "CN2" , "CN9" , "CN17"),
              #minprop = 0.3,
              progressbar = TRUE
)
res.cat <- surv_categorize(a)
res.cat$cluster <- mc_steele_active_4$classification[rownames(res.cat)]

fit <- survfit(Surv(OS.time, OS) ~ cluster, data = res.cat)
steele_OS<-ggsurvplot(fit, data = res.cat, risk.table = TRUE, conf.int = FALSE, pval = TRUE, palette  = c("#72F59A", "#F57D72","#F7C073","#8672F5"), censor = FALSE)

a<-surv_cutpoint(steele_f,
              time = "PFI.time",
              event = "PFI",
              variables = c("CN1" , "CN2" , "CN9" , "CN17"),
              #minprop = 0.3,
              progressbar = TRUE
)
res.cat <- surv_categorize(a)
res.cat$cluster <- mc_steele_active_4$classification[rownames(res.cat)]

fit <- survfit(Surv(PFI.time, PFI) ~ cluster, data = res.cat)
steele_PFI<-ggsurvplot(fit, data = res.cat, risk.table = TRUE, conf.int = FALSE, pval = TRUE, palette  = c("#72F59A", "#F57D72","#F7C073","#8672F5"), censor = FALSE)
```

## Tao
```{r}
a<-surv_cutpoint(tao_f,
              time = "OS.time",
              event = "OS",
              variables = c("Sig1" , "Sig2" , "Sig5" , "Sig7"),
              progressbar = TRUE
)
res.cat <- surv_categorize(a)
res.cat$cluster <- mc_tao_active_4$classification[rownames(res.cat)]

fit <- survfit(Surv(OS.time, OS) ~ cluster, data = res.cat)
tao_OS<-ggsurvplot(fit, data = res.cat, risk.table = TRUE, conf.int = FALSE, pval = TRUE, palette  = c("#72F59A","#F57D72","#F7C073"), censor = FALSE)

a<-surv_cutpoint(tao_f,
              time = "PFI.time",
              event = "PFI",
              variables = c("Sig1" , "Sig2" , "Sig5" , "Sig7"),
              #minprop = 0.3,
              progressbar = TRUE
)
res.cat <- surv_categorize(a)
res.cat$cluster <- mc_tao_active_4$classification[rownames(res.cat)]

fit <- survfit(Surv(PFI.time, PFI) ~ cluster, data = res.cat)
tao_PFI<-ggsurvplot(fit, data = res.cat, risk.table = TRUE, conf.int = FALSE, pval = TRUE, palette  = c("#72F59A","#F57D72","#F7C073"), censor = FALSE)
```

# Figure 6
```{r}
drew_PFI$plot <- drew_PFI$plot + labs(x = NULL, y = NULL)
drew_OS$plot <- drew_OS$plot + labs(x = NULL) 

steele_OS$plot <- steele_OS$plot + labs(x = NULL)
steele_PFI$plot <- steele_PFI$plot + labs(y = NULL, x = NULL) 

tao_PFI$plot <- tao_PFI$plot + theme(legend.position = "none")  + labs( y = NULL) 

surv_list_plot<- list(drew_OS$plot, drew_PFI$plot,
                 steele_OS$plot, steele_PFI$plot, 
                 tao_OS$plot, tao_PFI$plot)
```

# Supplementary risk tables
```{r}
surv_list_plot<- list(drew_OS$table, drew_PFI$table,
                 steele_OS$table, steele_PFI$table, 
                 tao_OS$table, tao_PFI$table)
```

# Stratified Cox analysis
```{r stratified-cox-function}
run_stratified_cox <- function(df, cluster_vec, sig_vars, time_var, event_var, label) {

  a <- surv_cutpoint(df, time = time_var, event = event_var,
                     variables = sig_vars, progressbar = FALSE)
  res.cat <- surv_categorize(a)
  res.cat$cluster      <- cluster_vec[rownames(res.cat)]
  res.cat$cancer_type  <- df$cancer_type

  # intersect with purity table
  tt    <- intersect(rownames(res.cat), rownames(purezza))
  prova <- cbind(res.cat[tt, , drop = FALSE],
                 purezza[tt,  , drop = FALSE],
                 df[tt, 1, drop = FALSE])

  cox <- coxph(
    as.formula(paste0("Surv(", time_var, ", ", event_var,
                      ") ~ as.factor(cluster) + age + purity + strata(cancer_type)")),
    data = prova
  )

  tbl <- cox %>%
    tbl_regression(exp = TRUE, include = "as.factor(cluster)") %>%
    bold_p(t = 0.05) %>%
    modify_caption(paste0("**", label, "** – cluster HR (stratified by tumor type)"))

  list(cox = cox, tbl = tbl, data = prova)
}
```

```{r stratified-cox-models}
drew_strat_OS  <- run_stratified_cox(drew_f,  mc_drew_active_4$classification,
                                     c("CX1","CX2","CX3","CX5"),
                                     "OS.time",  "OS",  "Drew – OS")

drew_strat_PFI <- run_stratified_cox(drew_f,  mc_drew_active_4$classification,
                                     c("CX1","CX2","CX3","CX5"),
                                     "PFI.time", "PFI", "Drew – PFI")

steele_strat_OS  <- run_stratified_cox(steele_f, mc_steele_active_4$classification,
                                       c("CN1","CN2","CN9","CN17"),
                                       "OS.time",  "OS",  "Steele – OS")

steele_strat_PFI <- run_stratified_cox(steele_f, mc_steele_active_4$classification,
                                       c("CN1","CN2","CN9","CN17"),
                                       "PFI.time", "PFI", "Steele – PFI")

tao_strat_OS  <- run_stratified_cox(tao_f, mc_tao_active_4$classification,
                                    c("Sig1","Sig2","Sig5","Sig7"),
                                    "OS.time",  "OS",  "Tao – OS")

tao_strat_PFI <- run_stratified_cox(tao_f, mc_tao_active_4$classification,
                                    c("Sig1","Sig2","Sig5","Sig7"),
                                    "PFI.time", "PFI", "Tao – PFI")

drew_strat_OS$tbl
drew_strat_PFI$tbl
steele_strat_OS$tbl
steele_strat_PFI$tbl
tao_strat_OS$tbl
tao_strat_PFI$tbl
```

# Figure 7
```{r forest-plot-cluster2}
# Adjust all cluster coefficients from the six Cox models as one family.
cox_models <- list(
  Drews_OS   = drew_strat_OS$cox,
  Drews_PFI  = drew_strat_PFI$cox,
  Steele_OS  = steele_strat_OS$cox,
  Steele_PFI = steele_strat_PFI$cox,
  Tao_OS     = tao_strat_OS$cox,
  Tao_PFI    = tao_strat_PFI$cox
)

all_cluster_tests <- do.call(
  rbind,
  lapply(names(cox_models), function(model_id) {
    coefficient_table <- summary(cox_models[[model_id]])$coefficients
    cluster_terms <- startsWith(
      rownames(coefficient_table),
      "as.factor(cluster)"
    )
    data.frame(
      model_id = model_id,
      term = rownames(coefficient_table)[cluster_terms],
      p_raw = coefficient_table[cluster_terms, "Pr(>|z|)"],
      row.names = NULL
    )
  })
)
all_cluster_tests$p_holm <- p.adjust(
  all_cluster_tests$p_raw,
  method = "holm"
)

# Forest plot: Cluster 2 vs Cluster 1 HR across all datasets/endpoints.
cluster2_tests <- all_cluster_tests[
  all_cluster_tests$term == "as.factor(cluster)2",
  c("model_id", "p_holm")
]

forest_data <- data.frame(
  model_id = c("Drews_OS", "Drews_PFI", "Steele_OS",
               "Steele_PFI", "Tao_OS", "Tao_PFI"),
  dataset  = c("Drews", "Drews", "Steele", "Steele", "Tao", "Tao"),
  endpoint = c("OS", "PFI", "OS", "PFI", "OS", "PFI"),
  HR       = c(0.64, 0.68, 0.85, 0.85, 0.92, 0.87),
  CI_low   = c(0.54, 0.58, 0.76, 0.77, 0.84, 0.79),
  CI_high  = c(0.77, 0.80, 0.95, 0.94, 1.01, 0.95)
)
forest_data$p_holm <- cluster2_tests$p_holm[
  match(forest_data$model_id, cluster2_tests$model_id)
]
forest_data$sig <- forest_data$p_holm < 0.05

# Label: dataset + endpoint, ordered for display
forest_data$label <- factor(
  paste0(forest_data$dataset, " – ", forest_data$endpoint),
  levels = rev(c("Drews – OS", "Drews – PFI",
                 "Steele – OS", "Steele – PFI",
                 "Tao – OS", "Tao – PFI"))
)

forest_data$p_label <- ifelse(
  forest_data$p_holm < 0.001,
  "Holm P < 0.001",
  paste0("Holm P = ", sprintf("%.3f", forest_data$p_holm))
)

figure7 <- ggplot(forest_data, aes(x = HR, y = label)) +
  # reference line
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  # CI bars
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high, color = sig),
                 height = 0.25, linewidth = 0.8) +
  # HR point
  geom_point(aes(color = sig, shape = sig), size = 3.5) +
  # p-value labels on right
  geom_text(aes(x = 1.04, label = p_label), hjust = 0,
            size = 3.2, color = "grey30") +
  scale_color_manual(values = c("TRUE" = "#2166AC", "FALSE" = "#999999"),
                     labels = c("TRUE" = "Holm P < 0.05",
                                "FALSE" = "Holm P ≥ 0.05")) +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1),
                     labels = c("TRUE" = "Holm P < 0.05",
                                "FALSE" = "Holm P ≥ 0.05")) +
  # dataset separator lines
  geom_hline(yintercept = c(2.5, 4.5), linetype = "dotted", color = "grey70") +
  # dataset labels on left margin
  annotate("text", x = 0.47, y = 6.3, label = "Drews",   fontface = "bold", size = 3.5, hjust = 0) +
  annotate("text", x = 0.47, y = 4.3, label = "Steele", fontface = "bold", size = 3.5, hjust = 0) +
  annotate("text", x = 0.47, y = 2.3, label = "Tao",    fontface = "bold", size = 3.5, hjust = 0) +
  scale_x_continuous(
    limits = c(0.45, 1.30),
    breaks = c(0.5, 0.6, 0.7, 0.8, 0.9, 1.0)
  ) +
  labs(
    title    = "Cluster 2 vs Cluster 1 – stratified Cox (adjusted for tumor type, age, purity)",
    x        = "Hazard Ratio (95% CI)",
    y        = NULL,
    color    = NULL, shape = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position  = "bottom",
    panel.grid.major.x = element_line(color = "grey90"),
    plot.title = element_text(size = 11)
  )

figure7
```